Inference using context-dependent models of mutation

### 2021-10-05 writeups/context-dependence.tex
More comments

As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
to change $x$ into $y$, i.e.
\[
J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}.
\]
The rate $G(n)_{x,y}$ is then the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
    G(n)_{x,y} = \sum_{i=1}^n \phi(E(y)-E(x)) \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the rates of all transition triples
that would change $x$ into $y$.
In principle, this gives us the transition probabilities for the process by a matrix exponential:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
\end{align}
which would then provide a route to parameter estimation.
In practice, the size of these matrices makes direct application obviously impractical.
This paper proposes an alternate strategy.

@@

\begin{figure}
    \begin{center}
        \includegraphics{Tmer-dependency}
    \end{center}
    \caption{
        \textbf{(A)} A $(6,2,2)$ T-mer.
        Nonmutated sites are shown in lower case letters.
        \textbf{(B)}
        The three mutations have extended the influence of the circled \nA{}
        in the initial sequence (top)
        to the six positions colored red in the final sequence.
        Red dotted arrows show the propagation of dependency
        between the circled \nA{}
        and the third site in the final sequence (bottom).
        The maximum window width, $w$, is equal to 1.
        \label{fig:Tmers}
    }
\end{figure}

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
and that something will be sufficient to compute the full likelihood.
To do this, let $p_{n,\ell,r}(x,y)$ for $r+\ell < n$ denote
the probability that an initial sequence $x$ of length $n$, after time $t$,
is found to match a smaller sequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
We refer to such patterns as ``T-mers'', as depicted in Figure \ref{fig:Tmers}.
This quantity is found by marginalizing $p_m$:
\begin{align}
    p_{n,\ell,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
We expect $p_{n,\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
\begin{align} \label{eqn:window_approx}
    p_{n,\ell,r}(x,y) \approx \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}.
\end{align}

@@

What we actually want is the \emph{full likelihood} $\like(y|x)$,
the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
This can be computed as the product of the probability of each site in $y$ given $x$ and previous sites:
defining $x_{<i}=(x_1, \ldots, x_{i-1})$, this is:
\begin{align*}
    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} .
\end{align*}
Using the same methods, we can approximate the terms in this sum
by discarding approximately independent sites from the conditioning.
%EM The following sentence may get us in trouble with the formalization police.
If $\ell$ is long enough that approximation \eqref{eqn:window_approx} holds, then:
%EM This casual reader is a little confused about the left to right nature here.
\begin{align} \label{eqn:full_approx}
    \P\{ Y_i = y_i \given Y_{<i} \;\&\; X \}
    &\approx
        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\&\; X_{i-3\ell}^{(4\ell+1)} \}  \\
    &=
        \frac{
            p_{2\ell+1,\ell,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
        }{
            p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
        } .
\end{align}
By rearranging the product,
this gives us an expression for the full likelihood as the ratio of the likelihoods of all $(n,\ell,\ell)$ T-mers in $(x,y)$
to that of of all $(n,\ell,\ell+1)$ T-mers.
% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
% matched to pattern $b$ in $y$ at a position offset by $\ell$.
% Then
% \begin{align}
%     \log \like(y|x)
%     &\approx
%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
%         -
%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
% \end{align}

@@

Therefore, the sequence at a particular site may directly affect the process at the neighboring $w$ sites on either side.
If another change happens, say, at the end of this window,
then the indirect effects of this change may extend out to $2w-1$ sites.
In this way, the influence of the sequence at each site extends outwards, as shown in Figure \ref{fig:Tmers} --
so if we take the overhangs long enough that this dependency is unlikely to extend to the base of the Tmer,
the approximation will be good.
As we show in Section \ref{ss:approx_pf}, if the overhangs are $\ell = r = kw$,
then the error in approximation \eqref{eqn:window_approx}
is bounded by the probability that there are more than $k$ changes in a given window of length $w$.
Concretely, sites further away than $kw$ only matter if there is a chain of at least $k$ intervening mutations
-- which is unlikely if $k$ is larger than the expected number.

>>>>

As noted above, there may be more than one way to mutate a such a sequence $x$ to get such a $y$:
for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
to change $x$ into $y$, i.e.
\[
J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}.
\]
The rate $G(n)_{x,y}$ is then the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
    G(n)_{x,y} = \sum_{i=1}^n \phi(E(y)-E(x)) \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the rates of all transition triples
that would change $x$ into $y$.
In principle, this gives us the transition probabilities for the process by a matrix exponential:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
\end{align}
which would then provide a route to parameter estimation.
In practice, the size of these matrices makes direct application obviously impractical.
This paper proposes an alternate strategy.

@@

\begin{figure}
    \begin{center}
        \includegraphics{Tmer-dependency}
    \end{center}
    \caption{
        \textbf{(A)} A $(6,2,2)$ T-mer.
        Nonmutated sites are shown in lower case letters.
        \textbf{(B)}
        The three mutations have extended the influence of the circled \nA{}
        in the initial sequence (top)
        to the six positions colored red in the final sequence.
        Red dotted arrows show the propagation of dependency
        between the circled \nA{}
        and the third site in the final sequence (bottom).
        %EM This "window" terminology... should we have another quick definition here?
        The maximum window width, $w$, is equal to 1.
        \label{fig:Tmers}
    }
\end{figure}

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
%EM The full exact likelihood? Isn't it an approximation?
and that something will be sufficient to compute the full likelihood.
To do this, let $p_{n,\ell,r}(x,y)$ for $r+\ell < n$ denote
%EM Can we say initial subsequence or something?
the probability that an initial sequence $x$ of length $n$, after time $t$,
is found to match a smaller sequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
%EM Say $a$ is the first $\ell$ bases of $x$ and $b$ is the last $r$ bases of $x$.
We refer to such patterns as ``T-mers'', as depicted in Figure~\ref{fig:Tmers}.
%EM Later we use the terminology "base of the T-mer." Let's define that here.
%EM Shall we define w here?
This quantity is found by marginalizing $p_m$:
\begin{align}
%EM Are we parameterizing in terms of n,\ell,r or n,\ell,m? We see both.
    p_{n,\ell,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
We expect $p_{n,\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
%EM should this be p_{n,\ell,m} below?
%EM What is k here? And do you mean \ell?
is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
\begin{align} \label{eqn:window_approx}
%EM I think I know what you're doing here but the $i$ seems rather lonely here. Do we think of it as something for marginalization or summation or something?
%EM Also, x on the left hand side appears to be the full sequence, while on the right hand side it appears to be a subsequence. Same for y.
    p_{n,\ell,r}(x,y) \approx \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}.
\end{align}

@@

What we actually want is the \emph{full likelihood} $\like(y|x)$,
the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
This can be computed as the product of the probability of each site in $y$ given $x$ and previous sites:
defining $x_{<i}=(x_1, \ldots, x_{i-1})$, this is:
\begin{align*}
%EM I'm not accustomed to seeing a big chunky ampersand in prob statements, rather I typically see commas. But you be you if this is how you do it.
    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} .
\end{align*}
Using the same methods, we can approximate the terms in this sum
by discarding approximately independent sites from the conditioning.
%EM The following sentence may get us in trouble with the formalization police.
If $\ell$ is long enough that approximation \eqref{eqn:window_approx} holds, then:
\begin{align} \label{eqn:full_approx}
    \P\{ Y_i = y_i \given Y_{<i} \;\&\; X \}
    &\approx
        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\&\; X_{i-3\ell}^{(4\ell+1)} \}  \\
    &=
        \frac{
            p_{2\ell+1,\ell,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
        }{
            p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
        } .
\end{align}
By rearranging the product,
this gives us an expression for the full likelihood as the ratio of the likelihoods of all $(n,\ell,\ell)$ T-mers in $(x,y)$
to that of of all $(n,\ell,\ell+1)$ T-mers.
% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
% matched to pattern $b$ in $y$ at a position offset by $\ell$.
% Then
% \begin{align}
%     \log \like(y|x)
%     &\approx
%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
%         -
%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
% \end{align}

@@

Therefore, the sequence at a particular site may directly affect the process at the neighboring $w$ sites on either side.
If another change happens, say, at the end of this window,
then the indirect effects of this change may extend out to $2w-1$ sites.
In this way, the influence of the sequence at each site extends outwards, as shown in Figure~\ref{fig:Tmers} --
so if we take the overhangs long enough that this dependency is unlikely to extend to the base of the Tmer,
the approximation will be good.
As we show in Section \ref{ss:approx_pf}, if the overhangs are $\ell = r = kw$,
then the error in approximation \eqref{eqn:window_approx}
is bounded by the probability that there are more than $k$ changes in a given window of length $w$.
Concretely, sites further away than $kw$ only matter if there is a chain of at least $k$ intervening mutations
-- which is unlikely if $k$ is larger than the expected number.
